Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2024-12-05 index.qmd
section:issues,CaseV; process:write

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its most general form, Thurstone's theory deals with pairwise comparisons of stimuli made by a single judge [@Thurstone_1927b, pp. 267]. The theory proposes that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* represents the psychological impact each stimulus has on judges or, more simply, their underlying perception of the stimulus' trait level. According to the theory, the discriminal process for each stimulus follows a Normal distribution, where its mode (mean), referred to as the *modal discriminal process*, indicates the stimulus' position on the trait continuum, and its dispersion, known as the *discriminal dispersion*, reflects the perceived trait variability of the stimulus.

@@

<!-- 3. Thurstone's Theory of CJ: discriminal difference -->
However, because the discriminal process of a single stimulus is not directly observable, the theory introduces the *law of comparative judgment*. This law posits that in pairwise comparisons, a judge perceives the stimulus positioned further along the trait continuum as having a higher level of that trait. This principle highlights that the outcome of a pairwise comparison likely depends on the relative distance between stimuli rather than their absolute positions on the trait continuum.

@@

<!-- 4. Example of discriminal difference -->
@fig-discriminal_difference shows the distribution of the discriminal difference for the texts depicted in @fig-discriminal_process, assuming a correlation of $\rho = 0.6$. The figure reveals that, under these conditions, the judge perceives text B as having significantly higher quality than text A, as indicated by the shaded gray area under the curve $P(B > A)$. As a result, the dichotomous outcome of this comparison would mostly favor text B over text A.

@@

<!-- 6. example of the effects of correlation on the discriminal difference -->
@fig-correlation illustrates how varying correlations influence the distribution of the discriminal difference for the texts depicted in @fig-discriminal_process. Since the texts differ in quality, higher correlations increase the likelihood that the discriminal difference distinctly favors text B over text A. This is evident from the larger proportion of the area under the curve, $P(B > A)$, that lies above zero (shaded gray area in @fig-correlation) compared to curves with lower correlations. Moreover, although the figure does not illustrate this scenario, it is reasonable to infer that if the texts had similar or identical quality levels, higher correlations would likely reduce the chance of the discriminal difference distinctly favoring one text over the other. This probability reduction occurs because the distribution of the discriminal difference would become more narrowly centered around zero.

@@

<!-- 4. What does the equal dispersion between stimuli imply? and does it hold? -->
On the one hand, the assumption of *equal dispersion between stimuli* suggests that the perceived trait variability is consistent across all stimuli. However, Thurstone observed that this assumption may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376], as these traits and stimuli can introduce judgment discrepancies due to their unique characteristics. Research on text quality may already provide evidence of this assumption's violation, particularly when considering the role of misfit statistics in the analysis of CJ data [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018]. Misfit statistics quantify the judgment discrepancies associated with a given stimulus [@Pollitt_2004, pp. 12; @Goossens_et_al_2018, pp. 20]. For instance, texts identified as "misfits" usually exhibit more judgment discrepancies than others. This, in turn, implies that the discriminal difference associated with these texts shows a broader dispersion, ultimately suggesting that their discriminal processes also exhibit more variation than other texts. Notably, this reasoning also applies to "misfit" judges, whose evaluations reflect substantial deviations due to the unique characteristics of the stimuli or the judges themselves.

@@

<!-- 5. But what are the challenges when these assumptions are violated? -->
Moreover, assuming equal dispersion between stimuli despite its violation causes Case V (and the BTL model) to overlook critical differences in the reliability of the trait across stimuli. Moreover, if researchers recognize that misfit statistics help to identify such violations, the usual practice in the CJ literature of excluding stimuli based on these statistics [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018] risks discarding valuable information and introducing bias into the trait's estimates [@Zimmerman_1994; @McElreath_2020]. The direction and magnitude of these biases remain unpredictable because they depend on which stimuli researchers exclude from the analysis. Together, these oversights undermine the reliability of the trait and ultimately compromise its validity [@Perron_et_al_2015, pp. 2].

@@

<!-- 11. How to solve the violation of the assumption? -->
Fortunately, the same literature offers solutions for addressing these issues. @Andrich_1978 and @Wainer_et_al_1978 recommend integrating judges' biases into the BTL model. Moreover, the literature advocates for the incorporation of relevant hierarchical structures into the statistical model to account for these dependencies. Together, these additions can result in a model resembling a Multilevel Structural Equation Model (MSEM) [@Hoyle_et_al_2023, chap. 26] combined with a multidimensional or two-parameter logistic IRT model [@Hoyle_et_al_2023, chap. 15], depending on the theoretical and statistical treatment of judges' biases.

@@

<!-- 1. The BTL model as a measurement model for estimating latent traits -->
Building on the previous section, it is evident that the BTL model commonly functions as the trait's measurement model in CJ experiments [@Andrich_1978; @Bramley_2008]. A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010]. For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

<!-- 4. So, how do we mitigate these risks? -->
To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp. 138] and Item Response Theory (IRT) [@Fox_2010, chap. 6; @vanderLinden_et_al_2017_I, chap. 24] recommend conducting these analyses and tests within a structural model. A structural model specifies how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010]. This approach allows analyses that can account for both the BTL scores and their uncertainties simultaneously, rather than treating them as separate elements. Therefore, an integrated approach that combines CJ’s measurement and structural models can offer significant advantages.

>>>>

<!-- 1. Thurstone's Theory of CJ: discriminal process -->
In its most general form, Thurstone's theory deals with pairwise comparisons of stimuli made by a single judge [@Thurstone_1927b, pp. 267]. The theory proposes that two key factors determine the dichotomous outcome of these comparisons: the discriminal process of each stimulus and their discriminal difference. The *discriminal process* represents the psychological impact each stimulus has on judges or, more simply, their underlying perception of the stimulus' trait. According to the theory, the discriminal process for each stimulus follows a Normal distribution, where its mode (mean), referred to as the *modal discriminal process*, indicates the stimulus' position on the trait continuum, and its dispersion, known as the *discriminal dispersion*, reflects the variability of the stimulus' perceived trait.

@@

<!-- 3. Thurstone's Theory of CJ: discriminal difference -->
However, because the discriminal process of a single stimulus is not directly observable, the theory introduces the *law of comparative judgment*. This law posits that in pairwise comparisons, a judge perceives the stimulus positioned further along the trait continuum as having a higher level of that trait. This emphasizes that the outcome of a pairwise comparison likely depends on the relative distance between stimuli rather than their absolute positions on the trait continuum.

@@

<!-- 4. Example of discriminal difference -->
@fig-discriminal_difference shows the distribution of the discriminal difference for the texts depicted in @fig-discriminal_process, assuming a correlation of $\rho = 0.6$. The figure reveals that, under these conditions, the judge perceives text B as having significantly higher quality than text A, as indicated by the shaded gray area under the curve $P(B > A)$. As a result, the dichotomous outcome of this comparison would likely favor text B over text A.

@@

<!-- 6. example of the effects of correlation on the discriminal difference -->
@fig-correlation illustrates how varying correlations influence the distribution of the discriminal difference for the texts depicted in @fig-discriminal_process. Since the texts differ in quality, higher correlations increase the likelihood that the discriminal difference distinctly favors text B over text A. This is evident from the larger proportion of the area under the curve, $P(B > A)$, that lies above zero compared to curves with lower correlations. Moreover, although the figure does not illustrate this scenario, it is reasonable to infer that if the texts had similar or identical quality levels, higher correlations would likely reduce the chance of the discriminal difference distinctly favoring one text over the other. This probability reduction occurs because the distribution of the discriminal difference would become more narrowly centered around zero.

@@

<!-- 4. What does the equal dispersion between stimuli imply? and does it hold? -->
On the one hand, the assumption of *equal dispersion between stimuli* suggests that the variability of the perceived trait remains consistent across all stimuli. However, Thurstone contended that this assumption may not hold when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927a, pp. 376], as these traits and stimuli can introduce judgment discrepancies due to their unique features. Indeed, evidence of such violation may already exist in the CJ literature as misfit statistics. These statistics measure the judgment discrepancies associated with a given stimulus [@Pollitt_2004, pp. 12; @Goossens_et_al_2018, pp. 20]. For instance, labeling texts as "misfits" indicates that comparisons involving these texts result in more judgment discrepancies than comparisons involving other texts [@Pollitt_2012a; @Pollitt_2012b; @vanDaal_et_al_2016; @Goossens_et_al_2018]. This finding implies that the discriminal differences associated with "misfits" texts usually display a broader dispersion, suggesting that their discriminal processes also exhibit more variation than other texts. Notably, this reasoning also applies to "misfit" judges, whose evaluations reflect substantial deviations from the shared consensus due to the unique characteristics of the stimuli or the judges themselves.

@@

<!-- 5. But what are the challenges when these assumptions are violated? -->
However, assuming equal dispersions between stimuli despite its violation can lead Case V (and the BTL model) to overlook critical differences in the reliability of the trait across stimuli. Furthermore, if researchers acknowledge that misfit statistics help identify these critical differences, the common practice in the CJ literature of excluding stimuli based on these statistics [@Pollitt_2012b; @vanDaal_et_al_2017; @Goossens_et_al_2018] risks discarding valuable information and introducing bias into the trait's estimates [@Zimmerman_1994; @McElreath_2020, chap. 12]. The direction and magnitude of these biases are unpredictable, as they depend on the particular stimuli researchers exclude from the analysis. Taken together, these oversights undermine the reliability of the trait and ultimately compromise its validity [@Perron_et_al_2015, pp. 2].

@@

<!-- 11. How to solve the violation of the assumption? -->
Fortunately, the same literature offers solutions for addressing these issues. @Andrich_1978 and @Wainer_et_al_1978 recommend integrating judges' biases into the BTL model. Moreover, the literature advocates for the incorporation of relevant hierarchical structures into the statistical model to account for these dependencies. Together, these additions can result in a model resembling a Multilevel Structural Equation Model (MSEM) [@Hoyle_et_al_2023, chap. 26] combined with a multidimensional or two-parameter logistic Item Response Theory (IRT) model [@Hoyle_et_al_2023, chap. 15], depending on the theoretical and statistical treatment of judges' biases.

@@

<!-- 1. The BTL model as a measurement model for estimating latent traits -->
Building on the previous section, it is evident that, besides its inconveniences, the BTL model commonly functions as the trait's measurement model for CJ data [@Andrich_1978; @Bramley_2008]. A measurement model specifies how manifest variables contribute to the estimation of latent variables [@Everitt_et_al_2010]. For example, when evaluating text quality, researchers use the BTL model to process the dichotomous outcomes resulting from the pairwise comparisons (the manifest variables) to estimate scores that reflect the underlying quality level of texts (the latent variable) [@Laming_2004; @Pollitt_2012b; @Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023].

@@

<!-- 4. So, how do we mitigate these risks? -->
To mitigate these risks, principles from Structural Equation Modeling (SEM) [@Hoyle_et_al_2023, pp. 138] and Item Response Theory (IRT) [@Fox_2010, chap. 6; @vanderLinden_et_al_2017_I, chap. 24] recommend conducting these analyses and tests within a structural model. A structural model specifies how different manifest or latent variables influence the latent variable of interest [@Everitt_et_al_2010]. This approach allows analyses that can account for both the BTL scores and their uncertainties simultaneously, rather than treating them as separate elements.
